Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-10-22 analysis/paper/paper.qmd
editing 'Modeling the social context' section

To measure transmission biases in tool making, we draw on Darwinian evolutionary theory. This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural]. Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection. The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties. Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying]. We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.

@@

Although Garvey and Bettinger and Eerkens were unable to directly observe transmission processes because of their focus on archaeological data, Mesoudi and O'Brien [-@mesoudi2008cultural] conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results validate the robustness of previous assumptions about cultural transmission biases and their consequences for variation and correlation in material culture.

# Modeling the social context of the appearance of stemmed points {.unnumbered}

Inspired by the previous studies that used Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to understand the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum for foraging groups and how they started to make stemmed points: on one end we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement. This trial and error behavior leaves a distinctive signature on metric variables of the stemmed points. If trial and error was the dominant bias in the appearance of stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation between sites.

On the other end of our spectrum, socially connected groups occupied places close to other groups or had regular contact with other groups over long or short distances. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. The model stemmed point might be a highly successful and frequently chosen one, and the learners copy all information about the point design as a package. In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will be less varied.

>>>>

To measure transmission biases in tool-making, we draw on Darwinian evolutionary theory. This body of theory has helped archaeologists understand a variety of technological innovations and related human behaviors, for example using human behavioral ecology, phylogenetics, cultural transmission theory, and niche construction theory [@dunnell1980evolutionary; @richerson1992cultural; @lipo1997population; @bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural]. Cultural evolutionary approaches are effective at explaining cultural changes using mechanisms of inheritance, variation-generating processes, and selection. The mechanisms are similar to biological evolution, but in cultural systems they have unique and distinctive properties. Social transmission strongly affects these key evolutionary processes [@Whiten_2017], and can occur through various learning processes, such as stimulus enhancement, emulation, imitation, and teaching [@o2000applying; @schillinger2014copying]. We focus here on how different social contexts can result in different modes of social transmission of tool-making skills among hunter-gather populations.

@@

Due to their focus on archaeological data, Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

# Modeling the social context of the appearance of stemmed points in the Late Palaeolithic of Korea {.unnumbered}

Inspired by Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement. This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package. In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.
